---
title: "BRCmap - An R package to produce species distribution maps in styles suitable for use in distribution atlases"
author: "Colin Harrower"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BRCmap - An R package to produce species distribution maps in styles suitable for use in distribution atlases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Introduction 
BRCmap is an R package developed within the [Biological Record Centre (BRC)](http://www.brc.ac.uk) to simplify the production of publication quality distributions maps. BRCmap was initially designed for internal use within BRC but once developed it was felt it could be of use in the wider community, particularly recording schemes and the biological recording community in the UK.  As BRC typically deals with data from Britain, Ireland and the Channel Islands the package is focused upon producing distributions maps for these regions.

BRCmap includes functions and options that allow you to tailor the distribution map to your requirements as well as accessory functions that help with dealing with and or processing UK grid reference data. In summary the packages contains functions to:

- Interpret and convert the grid reference formats typically used within the Britain, Ireland and Channel Islands, specifically the Ordnance Survey of Great Britain (OSGB), the Ordnance Survey of Ireland (OSI), and the truncated version of the Miltary Grid Rerence System (MGRS) often used for the Channel Islands).
- Create distributions maps with background layers such UK outline, country boundaries, or Watsonian County boundaries (all included as datasets in the BRCmap package) or even user supplied outlines loaded to R as spatial objects
- Add distribution data directly from grid reference data onto the background layers (e.g. UK outline) in a range of styles (e.g. as grid square boundaries, or as points/plotting symbols)
- Produce R spatial objects such as `SpatialPolygonsDataFrame`,`SpatialPointsDataFrame`,`Simple features polygons object`, `simple features points object` from grid reference data including any associated attribute information. These R spatial objects which can then be used within R in mapping or Geographic Information Systems (GIS) workflows and or saved to a number of spatial data file types, such as ESRI shapefiles, GeoPackage, etc for use in other GIS software.
```{r, echo = FALSE, message = FALSE, warning = FALSE}
  # Load the library
  library(BRCmap)
  # Load sf but suppress startup
    suppressPackageStartupMessages(library(sf))
```

The example below shows how BRCmap can be used to quickly create a simple distribution map. 
```{r, fig.width=5, fig.height=6.5, tidy=TRUE, eval = TRUE}
  # Reduce margins
  par(mar=c(0,0,0,0)+0.1)
  # Create a base maps of the UK (using lower resolution outline)
  plot_GIS(UK_lowres)
  # Create a vector of grid references
  grs = c("SP50","SO29","SN03","NO19","NH22",
            "H40","R99","S00","W09","WA50",
            "WV37","WV64","HY33","HU55","HP71"
            )
  # Add some grid references to the map
  plotUK_gr(grs, col="red")
```

# Creating Distribution Maps

BRCmap is designed to help produce occurrence maps of the kinds typically found in species atlases by allowing the maps to be plotted directly from location information given as grid references as is often the case for Biolgoical Records data in the UK & Ireland. The maps produced by BRCmap typically have two main components the base map or background layer, which is used for context and is often something like the UK/Ireland outline(s) or country/Vice County boundaries, and then one or more plotting layer where occurrence data is plotted either by outlining specific grid squares or representing grid squares with symbols. I

## Plotting the Base map
Each of the functions in BRCmap that plot grid reference data are designed to add the occurrence data onto an existing base map, e.g. the UK outline, and do not by themselves open or create a plotting device. Instead you will first need to create a base plot onto which the grid reference data can subsequently be plotted, this can be done using the `plot_GIS` function. This function is passed the name of the spatial object you want to plot/use as the base map, this can either be one of the base layer spatial datasets included with BRCmap (see the section on customising the base map for more details) or a users supplied spatial object.

```{r, fig.width=5, fig.height=6.5}
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Plot UK coastline background map object using lower resolution version
    plot_GIS(UK_lowres)

```




## Plot grid references as outlines
The simplest way to plot grid reference data is where each grid references is represented by a square polygon outlining the boundaries of the square represented by the grid reference(s). This type of plotting is done in BRCmap using the `plotUK_gr` function. The colour used to fill the square and the colour of the border can be altered by providing appropriate values to `col` and `border` arguments respectively. These arguments accept the standard methods of specifying colour(s) in R e.g. named colours, rgb values, hex values, etc. As this function will plot the outline of the square represented by a grid references When plotting maps at the full UK scale you may want to avoid plotting fine resolution grid references, with grid refererences finer than 10km square often being  difficult to see on standard size plots.

```{r, fig.width=5, fig.height=6.5}
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Create a vector of grid references to be plotted
    gridrefs <- c("SP49","WV57","H40","NN29","NT20","TL97", "SP24NW")
  # Setup base map
    plot_GIS(UK_lowres)
  # Now add grid references to base map
    plotUK_gr(gridrefs, col="lightgreen")
```


## Customising the Base Map
BRCmap includes a few different commonmly used outlines that can be used as base layer(s) for any mapping. The base mapping options included in BRCmap are the:

1. coastlines/outlines of the UK, Ireland and the Channel Islands (`UK`)
2. Country boundaries for the UK, Ireland and the Channel Islands (`UK_countries`)
3. Watsonian Vice county boundaries for the UK, Ireland and the Channel Islands (`UK_VC`)

```{r, fig.width=5, fig.height=6.5, echo = FALSE,fig.cap = "'UK' contains coastline/outlines of UK, Ireland and the Channel Islands"}
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Plot UK coastline background map object
    plot_GIS(UK_lowres)
```

```{r, fig.width=5, fig.height=6.5, echo = FALSE, fig.cap = "'UK_countries' contains country boundaries for UK, Ireland and the Channel Islands"}
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Plot UK coastline background map object
    plot_GIS(UK_countries_lowres)
```

```{r, fig.width=5, fig.height=6.5, echo = FALSE, fig.cap = "'UK_VC' contains Watsonian Vice County boundaries for UK, Ireland and the Channel Islands"}
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Plot UK coastline background map object
    plot_GIS(UK_VC_lowres)
```

The base map outlines are stored as `sp` `SpatialPolygonsDataFrame` objects using the OSGB Coordinate References System (EPSG: 27700). Each consists of a different features with an attribute table so they can be subset if needed, for instance the `UK` object has seperate features for GB, Ireland, Isle of Man and the Channel Islands, while the `UK_countries` has features for England, Scotland, Wales, Isle of Man, Northern Ireland, Republic of Ireland and the Bailiwicks of Jersey and Guernesey. Each of the base mapping options is also available as a lower quality spatial object where the boundaries have been simplified that can be used in situations where where the background doesn't need to be as complex/detailed and or when there are size limitations on the final objects.


```{r, fig.width=5, fig.height=6.5 }
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Plot UK coastline background map object
    plot_GIS(UK)
  # Show attribute/data table for features in the object
    UK@data
  # Plot subset (only Ireland)
    plot_GIS(subset(UK_lowres, REGION == "Ireland"))

```

If you want to plot a particular smaller section of one of the basemaps but whether it isn't a seperate feature of the dataset, e.g. a small part of a region/countyyou can do this by providing `xlim` and `ylim` values as arguments to the `plot_GIS` function. For example if you wanted to plot the Sheltland Isles you could use the following

```{r, fig.width = 5, fig.height=6.5}
  par(mar=c(0,0,0,0)+0.1)
  plot_GIS(UK_countries[UK_countries$COUNTRY == "Scotland",], xlim = c(400000,500000), ylim = c(1100000,1250000), grid.div = 1000)
```




# Dealing with grid references
## Introduction to Grid Reference formats used for Biological Recording in the UK, Ireland & Channel Islands
There are three main grid references systems typically used in the UK, Ireland, and Channel Islands for Biological Recording. These are the Ordnance Survey of Great Britain (OSGB) national grid reference system which is used in UK, inlcuding Isle of Man, the Ordnance Survey of Ireland (OSI) national grid reference system used in Northern Ireland & the Republic of Ireland, and  the Military Grid References System (MGRS) used in the Channel Islands (though in many cases these MGRS grid references for the Channel Islands are tructated to exclude the initial zone code 30U which covers the entire region).

The grid references from all three systems look similar with grid references from all three systems typically starting with one or two letters, followed by an even number of digits. Grid references from UK & Channel Islands start with two letters, e.g. SP6489 or WV6347, with Channel Islands grid references being recognisable as they will start with WA or WA, while grid references from the Irish national grid start with a single letter, e.g H4107. Grid references from the Channel Islands can sometimes be presented as the full MGRS grid references, in which there will be zone identifier 30U at the start of the grid reference preceeding the WA or WV, e.g. 30UWV6347, however as the same zone applied for all of the region often the zone identifier may be omitted.

The starting letters of the grid references effectively specify a specific 100km x 100km square and the digits after specifying a location or area within this square. The digits specify position as distances from the south west/bottom left corner of the square in meters. The first half of the digits specify the horizontal distance, or Easting, from this point while the latter half specify the vertical distance, or Northing, and as such there should always be an even number of digits in the grid reference.

The location described can be of varying resolutions depending on the number of digits provided in the grid reference. For example a 10km square is specified by one pair of digits (one easting & one northing digit), a 1km square by 2 pairs of digits (two easting & two northing digits) and so on as shown in the table below. As grid references always refer to a square area the resoultion of a grid references is often described by the dimension of one side rather than the area, for instance a "10km grid square" refers to a 10km x 10km area (or 100km^2) and not to a grid square that covers an area of 10km^2. A few of the historically most commonly used grid square resolutions have their own names, for instance a Hectad is the name for a 10km x 10km grid square/grid references, while a monad is a 1km x 1km grid square/reference.

Occasionally you may encounter grid references that have a one or two character letter code after the digit pair. These special cases are referred to a Quadrants or Tetrads and are grid references used to represent 5km x 5km squares or a 2km x 2km squares respectively. Tetrad and quadrant grid references are essentially 10km grid references to which an additional letter code has been appended to indicate particular areas within the 10km square.

Grid Reference | Square Resolution | digit pairs | Square Type
---------------|-------------------|-------------|------------
SU68           |  10km x 10km      |  1          |Hectad (10km)
SU68NW         |  5km x 5km        |  1          |Quadrant (5km)
SU68E          |  2km x 2km        |  1          |Tetrad (2km)
SU6189         |  1km x 1km        |  2          |Monad (1km)
SU616895       |  100m x 100m      |  3          |100m
SU61698959     |  10m x 10m        |  4          |10m
SU6169089594   |  1m x 1m          |  5          |1m



In a quadrant grid reference the 10km squares has been divided into four 5km x 5km squares, arranged in a 2 by 2 grid, with each sections being allocated a two letter code to represent the corner of 10km square that the relevant section occupies, e.g. SW, NW, SE, NE for South West, North West, South East, and North East corners respectively as shown in the figure below.

```{r, fig.width=4, fig.asp = 1, tidy=TRUE, echo = FALSE, eval = TRUE, fig.cap = "Layout of quadrants within a 10km square"}
  # Reduce margins
    par(mar=c(0,0,0,0)+0.1)
  # Create 10km square as a spatial polygon so it can be plotted
    sq10km = gr2sf_poly("SU68")
  # Create all quadrants within this 10km as spatial polygons
    sq_quads = gr2sf_poly(paste0("SU68",c("SW","NW","SE","NE")))
  # Plot this square
    plot(st_geometry(sq10km), lwd = 3, border = 3)
  # add quadrants to plot
    plot(st_geometry(sq_quads),lwd = 1, add = TRUE)
  # Add label to each quadrants
    text(gr_let2num(paste0("SU68",c("SW","NW","SE","NE")), centre = TRUE), labels = c("SW","NW","SE","NE"),font = 2, cex = 0.9)
    rm(sq10km, sq_quads)
```

While in a tetrad grid reference the 10km squares has instread been divided into 25 2km x 2km squares, laid out in a 5 by 5 grid, each assigned a letter running from A, in the lower left, to Z, in the upper right, but omitting O to avoid confusion with zeros.


```{r, fig.width=4, fig.asp = 1, tidy=TRUE, echo = FALSE, eval = TRUE, fig.cap = "Layout of tetrads within a 10km square"}
  # Reduce margins
  par(mar=c(0,0,0,0)+0.1)
  # Create 10km square as a spatial polygon so it can be plotted
    sq10km = gr2sf_poly("SU68")
  # Create all tetrads within this 10km as spatial polygons
    sq_tets = gr2sf_poly(paste0("SU68",LETTERS[-15]))
  # Plot this square
    plot(st_geometry(sq10km), lwd = 3, border = 3)
  # add tetrads to plot
    plot(st_geometry(sq_tets),lwd = 1, add = TRUE)
  # Add label to each tetrad
    text(gr_let2num(paste0("SU68",LETTERS[-15]), centre = TRUE), labels = LETTERS[-15],font = 2, cex = 0.9)
    rm(sq10km, sq_tets)
    
```

Hopefully this should be as enough information for you to understand and deal with these grid references but further information on these grid references systems can be found on the following Wikipedia pages [British National Grid](https://en.wikipedia.org/wiki/Ordnance_Survey_National_Grid), [Irish Grid Reference system](https://en.wikipedia.org/wiki/Irish_grid_reference_system), [Military Grid Reference System](https://en.wikipedia.org/wiki/Military_Grid_Reference_System).

## Standardising Grid References
Although the essential components of a grid reference and their order are consistent the exact formatting can vary dramatically, for example sometimes spaces or hypens are used to seperate the individual components of the grid reference. In order for the functions in BRCmap to work consistently the grid references will need to be standardised to the format expected by the package. The standardised format is to use capital letters for the intial letter codes and any tetrad or quadrant and to have no spaces or other characters seperating the components of the grid reference. BRCmap contains a function `fmt_gridref` specifically for this purpose. This function will do some basic checks to also exclude/remove grid references that don't match the expected formats (e.g. that have uneven numbers of digits, too many letters a the start). 
```{r}
# Create a vector of grid references in a range of different formats
# including some bad grid references
  raw_grs = c("SP 43 27","W09","Wa50","sn0408","sN 02-02-", "SU06053919","SU687")
  # Use the fmt_gridref function to standardise these grid refs
  fmt_gridref(raw_grs)

```
In this case the function has remove the spaces and hypens from the grid references, converted the letters to upper case and has noticed that the last grid references has an uneven number of digits. **Note** this function only checks the basic structure of the grid references not whether the actual grid references is actually a valid grid reference or makes sense for the system (e.g. out of the typical plotting boundaries)
  
## Altering grid reference precision

In many cases you may want to take a grid reference or set of grid references and convert them to a lower resolution (e.g. determine what 1km or 10km square they occur in). The function `reformat_gr` can be used for this;

```{r}
  reformat_gr("SU6169089594")
```
by default `reformat_gr` will reformat to 10km square, but the output resolution can be changed via the 2nd argument (`prec_out`) by supplying a resolution value in metres.
```{r}
  # Output to 100m grid reference
  reformat_gr("SU6169089594",100)
```
The input grid reference argument and or the output precision arguments can accept a single value or be given vector(s) of grid references and or precision values to allow a series of grid references to be output to a either a single precision or each output to different precisions.

```{r}
  # Create a set of input grid references
   test_gr = c("SU6169089594","SU68","SP258597")
  # Reformat all to 10km
   reformat_gr(test_gr)
  # Reformat each to different precision
   reformat_gr(test_gr, c(100,10000,1000))
  # Now reformat all to 1km
   reformat_gr(test_gr,1000)
```

Notice that in the last example `reformat_gr` does not by default reformat a grid reference if the output resolution is finer than the resolution of the supplied grid reference. In such cases it instead returns a `NA` as was the case of the 2nd grid references in the example above. In rare situation you may want to still force these coarser resolution records to a finer resolution grid reference by padding with zeros whjich can be achieved by setting the `pad_gr` argument to `TRUE` from the default `FALSE`, this will however display a message warning about false precision in grid references to make it is clear what is happening and that in most cases this is not desired or advisable.

```{r}
  reformat_gr(test_gr,1000, pad_gr = TRUE)
```


## Styles

## Converting grid references to eastings & northings and vice versa
  
## Converting between grid references and WGS84 latitudes & longitudes

